---
title: "Classification Models"
output: html_document
date: "2025-02-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nnet)
library(ISLR2)
```


# 1. Introduction
<Introduce the models being used>

# 2. Data
\<Describe the data \>

```{r}
data = Default
str(data)


```


## 2.1 Visualizing the Data

### Distribution of Balance
<What does this figure mean?.

```{r}
ggplot(data, aes(x=balance , fill=default))+
  geom_histogram(bins = 30, alpha=0.7, position='identity')+
  labs(title = "Distribution of Balance by Default Status",
       x= "Balance",
       y= "Count")

```

### Distribution of Income


```{r}
ggplot(data, aes(x=income, fill = default))+
  geom_histogram(bins = 30, alpha = 0.7, position = 'identity')+
  labs(title = "Distribution of Income by Default Status",
       x= "Income",
       y= "Count")

```


```{r}
ggplot(data, aes(x=income, fill = student))+
  geom_histogram(bins = 30, alpha = 0.7, position = 'identity')+
  labs(title = "Distribution of Income by Default Status",
       x= "Income",
       y= "Count")

```

### Student Status by Default

```{r}
ggplot(data, aes(x=student, fill = default))+
  geom_bar(position = 'dodge')+
  labs(title = "Default Status by Student Status",
       x= "Students",
       y= "Count")

```

## 4. Logistic Regression
### Fitting the Model
<Describe logistic Regression>


```{r}
logit_model = glm(default ~ balance, data=data, family=binomial)
summary(logit_model)

```


```{r}
data$predicted_prob = predict(logit_model, type="response")
head(data)

```

### Evaluate Model Performance
<talk about model and Evaluation>
```{r}
threshold = 0.5
data$predicted_default = ifelse(data$predicted_prob > threshold, "Yes", "No")
conf_matrix = table(data$predicted_default, data$default)
conf_matrix

```


```{r}
accuracy = sum(diag(conf_matrix)) / sum(conf_matrix)
accuracy

```


# 5 Multiple Logistic Regression
## 5.1 Fitting the model
Here we will include an **interaction term** between 'income' and 'student' that will
allow the effect of 'income' on 'default' to differ between 'student' and 'non-student'

```{r}
logit_mult_model = glm(default ~ balance + income * student, data=data, family=binomial)
summary(logit_mult_model)

```

## 5.2 Evaluating the Model
<Talk about evaluation metrics/interpretation>

```{r}
data$mult_predicted_prob = predict(logit_mult_model, type = "response")
data$mult_predicted_default = ifelse(data$mult_predicted_prob > threshold, "Yes",
                                     "No")
conf_matrix_mult = table(data$predicted_default, data$default)
conf_matrix_mult

```


```{r}
accuracy_mult = sum(diag(conf_matrix_mult)) / sum(conf_matrix_mult)
accuracy_mult

```

# 6. Multinomial logistic Regression
## 6.1 Load the Data

```{r}
data2 = Carseats
data2$SalesCategory = cut(data2$Sales, breaks = 3, labels = c("Low", "Medium", "High"))

```


```{r}
multi_model = multinom(SalesCategory ~ Price + Income + Advertising, data = data2)
multi_model

```

##6.2 Make Predictions

```{r}
data2$nomial_predicted_SalesCat = predict(multi_model)
head(data2)

```

## 6.3 Evaluate Model

```{r}
conf_matrix_multi = table(data2$nomial_predicted_SalesCat, data2$SalesCategory)
conf_matrix_multi

```


```{r}
accuracy_multi = sum(diag(conf_matrix_multi)) / sum(conf_matrix_multi)
accuracy_multi

```