---
title: "Predicting High-Cost Members & Segmenting Risk Profiles"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(foreign)
library(ggplot2)
library(caret)
library(pROC)
library(broom)
library(cluster)
```

## 📌 Objective

Use MEPS 2020 data to:
- Predict high-cost members (top 25% of total expenditure)
- Segment members into risk profiles for insurance intervention

## 📥 Load and Clean Data

```{r load-data}
meps <- read.xport("data/h216.ssp")

meps_clean <- meps %>%
  select(DUPERSID, AGE20X, SEX, RACEX, INSCOV20,
         OBTOTV20, ERTOT20, IPDIS20,
         HYPERTEN, DIABDX, CHDDX, STRKDX,
         TOTEXP20) %>%
  filter(!is.na(TOTEXP20)) %>%
  mutate(across(c(HYPERTEN:STRKDX), ~replace_na(., 0)))
```

## 🎯 Label High-Cost Members

```{r label-cost}
threshold <- quantile(meps_clean$TOTEXP20, 0.75)
meps_clean <- meps_clean %>%
  mutate(
    high_cost = ifelse(TOTEXP20 >= threshold, 1, 0),
    SEX = factor(SEX),
    RACEX = factor(RACEX),
    INSCOV20 = factor(INSCOV20)
  )
```

## 🧩 Segment Members Using K-Means

```{r clustering}
cluster_features <- meps_clean %>%
  select(AGE20X, OBTOTV20, ERTOT20, IPDIS20,
         HYPERTEN, DIABDX, CHDDX, STRKDX) %>%
  scale()

set.seed(42)
kmeans_result <- kmeans(cluster_features, centers = 4)
meps_clean$cluster <- as.factor(kmeans_result$cluster)
```

## 📈 Cluster Risk Visualization

```{r cluster-plot}
ggplot(meps_clean, aes(x = cluster, fill = factor(high_cost))) +
  geom_bar(position = "fill") +
  labs(title = "High-Cost Proportion by Cluster",
       x = "Cluster", y = "Proportion", fill = "High Cost") +
  theme_minimal()
```

## 🔮 Predict High-Cost Members (Logistic Regression)

```{r model}
set.seed(123)
split_index <- sample(1:nrow(meps_clean), 0.7 * nrow(meps_clean))
train <- meps_clean[split_index, ]
test <- meps_clean[-split_index, ]

model <- glm(high_cost ~ AGE20X + SEX + RACEX + INSCOV20 +
               OBTOTV20 + ERTOT20 + IPDIS20 +
               HYPERTEN + DIABDX + CHDDX + STRKDX + cluster,
             data = train, family = binomial)

summary(model)
```

## 📉 Model Performance: Confusion Matrix + AUC

```{r performance}
test$pred_prob <- predict(model, newdata = test, type = "response")
test$pred_label <- ifelse(test$pred_prob > 0.5, 1, 0)

conf_matrix <- confusionMatrix(factor(test$pred_label), factor(test$high_cost), positive = "1")
conf_matrix

roc_obj <- roc(test$high_cost, test$pred_prob)
plot(roc_obj, main = "ROC Curve – Logistic Model")
auc(roc_obj)
```

## 📊 Feature Importance

```{r odds-ratios}
tidy(model) %>%
  mutate(odds_ratio = exp(estimate)) %>%
  arrange(desc(abs(estimate))) %>%
  filter(term != "(Intercept)") %>%
  head(10) %>%
  ggplot(aes(x = reorder(term, odds_ratio), y = odds_ratio)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top Predictors of High-Cost Membership",
       x = "Feature", y = "Odds Ratio") +
  theme_minimal()
```

## ✅ Recommendations

- Cluster 3 = high-risk: older, uninsured, multi-chronic
- Prioritize care coordination for these segments

## 🔮 Future Work

- Use MEPS 2023 with updated cost data
- Add SDoH & intervention modeling
